\chapter{Probability 101}
\label{chap:prob}

\vspace*{2cm}

\begin{center}
	\itshape
	"It is unanimously agreed that statistics
	depends somehow on probability. But, as to
	what probability is, and how it is connected
	with statistics, there has seldom been such
	complete disagreement and breakdown of
	communication since the Tower of Babel.
	
	
	Doubtless, much of the disagreement is
	merely terminological and would disappear
	under sufficiently sharp analysis."
	
	\vspace{0.5em}
	--- Savage, 1954, The foundations of statistics\todo{Add ref}
\end{center}

\vspace{2cm}


Throughout this course we will proceed with a physical interpretation of Probability, as we approach this complex field through the eyes of experimental physicists who want to \textit{measure something}. In nature, there exists some measurable quantities whose value depends (in some way) from unknown parameters. Observing these observables $x\in X$ we can make inferences on the nature of these unknown parameters $\mu\in A$. 

In this first treatment we have introduced the notation that will follow us for the rest of the course - observable quantities will be denoted with the Latin letters \textit{x, y, z}, while unknown parameters will be denoted with the Greek letters \textit{$\mu, \nu, \sigma$}. The important subtlety that this hopes to highlight is that these quantities live in different spaces, and that observables are quantities that we can measure while the parameters are not directly measurable, and the best we can hope to do is make an inference based on a measurement. Let's never forget this, as it will be a fundamental to not mix the two quantities in the future. 



\section{Operative definition of Probability}
\label{sec:prob_def}

We shall start with an operative definition of Probability, consistent with a physical quantity. Given an experimental procedure that is infinitely replicable, each repetition will correspond to an observable quantity $x\in X$. We shall call \textit{event} $E$ the subset of possible outcomes of $x$, $E\subset X$. The \textbf{probability of event E} in the frequentist sense is defined as:
	\[
	P(E) = \lim_{N\to\infty} \frac{n(E)}{N}
	\]

where $n(E)$ is the number of times that event $E$ is verified out of $N$ trials. In this operative definition, $P(E)$ is a measurable quantity, and thus it is fundamental that the experiment be repeatable infinite times in this formalism.

$X$ is the space of the possible outcomes of the observable $x$, and in the simple discrete example of a mix of colored balls inside of a box $X=\{ \text{red, blue, green}\}$ and $E$ can be any single extraction of a ball, or any combination. Each event defined this way will have its own probability, and as long as the experiment is infinitely repeatable this probability well defined and given by the frequentist (Von Mises) definition provided above. 


However, as mathematical limits are, for a limit to be well defined it must equal the same value in whichever way $N$ decides to reach infinity, no matter how the events are selected. This means that if I repeat an experiment to infinity, the probability of event $E$ must be the same if I consider all of the events, just the even ones, just the prime numbered ones, etc. 

\begin{example}
	If in the previous situation I extract (with replacement), the colored balls randomly to infinity, the sequence 
	\[
	\text{red, blue, green, red, blue, green, red, blue, green, red, }\dots
	\]
	
	does not have $P(\text{red})$ well defined, as if I consider all of the events in the limit I get $P(\text{red})=\frac{1}{3}$, but if I consider all of the events that follow the extraction of a green ball, $P(\text{red})=1$ and all those following a blue extraction $P(\text{red})=0$. These three limits were all taken in the mathematical sense to infinity, though they are not equal to the same value and therefore the limit is not well defined, and in fact this deterministic sequence does not have a probability defined. 
\end{example}

This means that frequentist probability is defined on an ensemble.\todo{This should be looked upon further, or omitted}


\subsection{Kolmogorov probability}
\label{subsec:kolmogorov}

	We define "probability" on the set $S$ some function of the set $P$\sidenote{Formally, $P$ is defined on a $\sigma$-algebra of subsets of $S$, so that we are sure that the union and conjugation operations remain within the same space} with the following properties:
\begin{enumerate}
	
	\item $\forall A \subset S \quad P(A)\geq 0$
	\item $P(S) = 1$
	\item $A_i \subset S_i ;\quad A_i\cap A_j = 0 \quad \forall i\neq j ;\quad \forall A_i \rightarrow P(\cup A_i) = \sum\limits_{i}P(A_i)$
	
\end{enumerate}

S can be taken as the space of all possible experimental observations, and each element a completely defined observation. In this case, $A$ represents a set of possible observations.


From these three axioms we can define a series of simple theorems\todo{Do we want to demonstrate these? Also maybe the formatting could do with some work}:


\begin{definition}[Conditional probability]
	Probability of event A given that event B has occurred
	\[
	P(A|B) \equiv \frac{P(A\cap B)}{P(B)}
	\]
This can be re-written in the more useful form:
	\[
	P(A\cap B)= P(A|B)\cdot P(B)
	\]
\end{definition}

$P(A|B)$ (also denoted with $P_B(A)$) is a Kolmogorov probability and satisfies all of the necessary axioms\todo{PROOF?}.

It is important to note that conditional probability has nothing to do with causation, correlation, or temporal relations between two events, simply put it is the probability of an event given another, and therefore is nothing more than a restriction of the probability space to a smaller ensemble. 

\begin{theorem}[Total probability law]
	Let $\{A_i\}$ be a countable partition of $S$ such that $\cup_i A_i = S$\todo{Here I omitted the intermediate passages for the proof which can be added in-line or as a proof later}. It follows that:
	\[
	\forall B: P(B)=\sum_i P(B\cap A_i) = \sum_i P(B|A_i)P(A_i)
	\]
\end{theorem}

This theorem is useful when the conditional probabilities are easier to calculate than the total.
\begin{example}
	I know that $25\%$ of lightbulbs are produced in Vietnam, while $75\%$ are produced in China. I know that of the Vietnamese lightbulbs, one in a thousand are broken during production while the Chinese ones are broken one percent of the time. I buy a lightbulb from the store. What is the probability that it is broken?
	
	The total probability law comes to help, as we can write:
	\begin{align*}
	P(\text{Broken}) &= P(\text{Broken}|\text{C})\cdot P(\text{C}) + P(\text{Broken}|\text{V})\cdot P(\text{V}) \\
	&= 0.01\times0.75 + 0.001\times0.25 \\
	&= 0.775\%
	\end{align*}
\end{example}

\begin{definition}[Independent events]
	Two events are independent if 
	\[
	P(A|B)=P(A)
	\]
	or, reformulated, two events are independent if
	\[
	P(A\cap B) = P(A)\cdot P(B)
	\]	
	This second form takes the name of the product law of probability
\end{definition}

It is important to remember that independence is not correlation, and the former only describes whether or not the probability of an event changes after the verification of another. For example, rolling a die and flipping a coin are two independent events, while picking two cards from a deck without replacement are not two independent events as removing one card from the deck changes the odds of picking the second. 

\begin{theorem}[Bayes' theorem]
	We have already written that 
	\[
	P(B\cap A) = P(B|A)\cdot P(A) \overset{\scriptscriptstyle \text{symm.}}{=} P(A|B)\cdot P(B)
	\]
	It follows naturally that
	\[
	P(B|A) = \frac{P(A|B)\cdot P(B)}{P(A)}
	\]
\end{theorem}	

The last expression takes the name of \textbf{Bayes' theorem}\sidenote{Interestingly enough, it is not what characterizes Bayesian statistics, and we remind that we are still in the Kolmogorov frequentist approach, though the theorem generalizes to the Bayesian approach if the probabilities are replaced with degrees of belief.}

\subsubsection{Spoiler: Bayesian probability}
\label{subsubsec:Bayes}

The Bayesian school uses a concept of probability different than the von Mises concept. To try to minimize the natural confusion that will follow this chapter, we will baptize this concept with a different name: \textit{"degree of belief"} $\pi(x)$. The idea of this subjective reasoning is that events of which we have similar information are by definition equally probable. This concept, differently than the frequentist definition we have studied already, can be applied to whatever quantity, even if the latter is not observable. It does not depend on the ensemble, but on the subject itself, and Bayesian d.o.b (degree of belief) has been shown to satisfy the Kolmogorov axioms. We will be returning on this concept in various chapters\todo{Maybe add which ones through a ref, and perhaps elaborate further once we understand better} and it will be crucial to not confuse it with the frequentist definition of probability. 

\subsection{Events as physical observables}
\label{subsec:observables}

In physics, we usually call "event" a measurement of a certain observable $X$ yielding the value $x\in O_X$, where $O_X$ is the space of all the possible results (values of $x$). We classify $X$ as a \textbf{random observable} if the probability of measuring each of its values (or set of values\sidenote{Let's never forget that the probability of measuring an exact value from a continuous distribution is exactly zero, and therefore for continuous observables we will always talk about the probability of measuring a set of values, for example $0\leq x\leq 1$.}) exists and is well defined. Physically, this implies that for any\sidenote{Note how this condition starts with "for all". It means that just because it is met for some subsets of the observable space it is not guaranteed that this probability is well defined for \textit{all} of them.} subset $A \subset O_X$, the frequentist limit converges:

	\[
	\forall A \subset O_X \quad \exists P(x\in A) = \lim_{N\to\infty}\frac{n(x\in A)}{N}
	\]


To move beyond empirical observation, math comes to the rescue, and we map this physical intuition to the formal mathematical framework of measure theory. This allows us to treat "measurements" as "functions" and inherit a powerful toolkit of existing theorems.

\begin{definition}[Random variable]
	A random variable is a function $X: S\rightarrow O_X$ such that:
	\[
	\forall A \subset O_X \quad X^{-1}(A)\subset \Sigma \rightarrow \exists P(X^{-1}(A))
	\]
\end{definition}


The function $X$  maps the quantity we have observed to a value that we will call "random variable". Since the outcome of an experiment can be a qualitative observable, or anything really, this bridge is helpful in order to import all of the nice properties that mathematics guarantees us. However, for this function to be mathematically sound, we must confirm that it is measurable. That means that for any numerical interval we are interested in (say, the probability of a value falling between 5 and 10) there must exist a corresponding, well-defined set of real-world outcomes ($X^{-1}(A)$) that sits within our "event space" ($\Sigma$). If this condition is met, we gain the power to assign a precise probability $P$ to those outcomes.

In essence, the random variable ensures that every "observable state of the world" captured by our instruments has a mathematically guaranteed seat at the table of probability theory.

This may all sound scary right now, but hopefully the following example will clarify that the concept, at its core, is really quite simple. 

\begin{example}
	Consider the simplest physical experiment: flipping a coin. The Sample Space $S$ contains two physical outcomes: $\{\text{Heads, Tails}\}$. To perform calculations, we define a Random Variable $X$ as a function that maps these physical states to numbers:
	
	\[
	X(s) = \begin{cases} 1 & \text{if } s = \text{Heads} \\ 0 & \text{if } s = \text{Tails} \end{cases}
	\]
	
	Here, the "Observable Space" $O_X$ is $\{0, 1\}$. If we want to find the probability of the set $A = \{1\}$ (the event "Heads"), we look at the pre-image:
	
	
	\[
	X^{-1}(\{1\}) = \{\text{Heads}\}
	\]
	
	Since "Heads" is a valid element of our event space $\Sigma$, the probability $P(X^{-1}(\{1\}))$ is well-defined (e.g., $0.5$). The function $X$ has successfully translated a physical piece of metal landing on a table into a number we can use in an equation.
\end{example}

Since we are looking to do \textit{something} with these random variables, it will surely be helpful to note that any function of a random variable is a random variable itself, and more importantly if a random variable assumes a probability distribution, each function of that variable will as well. Since we will be using "functions of random variables" quite often, let's give them a name: \textbf{statistics}.
\begin{definition}[Statistic]
	A statistic (singular) is a function of random variables $s(x,y,\dots)$, that depends only on the values themselves, not on unknown parameters. 
\end{definition}

The second part of this definition is important - a statistic only depends on \textit{measurable quantities} and is therefore also a measurable quantity. A nice, simple example of an important statistic is the mean of a set of data. This concept of statistic will follow us closely throughout the whole course, and while it's a good idea to introduce them now this will certainly not be the last time we update our definition.

\subsubsection{Probability distributions}
\label{subsubsec:pdf}

We have already thrown out this term \textit{probability distribution} in the previous section, and we won't get very far without giving it the proper definition it deserves. 

\begin{definition}[Probability mass function \textit{pmf}]
	For a discrete random variable X
	\[
	pmf(x)\equiv P(E:X=x)
	\]
\end{definition}

Often this gets (somewhat improperly) condensed in literature to be written as $P(X)$, which we will adopt from now on without too much hesitation. Probability mass functions are additive and satisfy our lovely Kolmogorov axioms\sidenote{They are defined positive and less than $1$, the sum of the total possibility space is $1$, and they are additive. The sum operation naturally changes definition based on the dimension of the outcome space.}, and so we can truly call them "probabilities". A simple example of a \textit{pmf} would be the case of the outcome of a fair six-sided die roll
	\[
	pmf(x) = \begin{cases} \frac{1}{6} & \text{if } x \in \{1,2,3,4,5,6\} \\ 0 &  \text{otherwise} \end{cases}
	\]

\newpage %Paragraph started on the old page, so the sidenote calculated on the wrong side

So much for discrete random variables, let's talk about the interesting stuff and generalize this formalism to continuous\sidenote{Naturally, we will never have a continuous result from an experiment as everything at some scale gets discretized or binned in some way, though we are sure the readers will have no problem with forgiving this abstraction.} observables. The probability mass function, as we have defined it, will be identically zero for each value in a continuous space such as $\Re$, so we will need a new definition, and much like we are used to in physics, the concept of \textit{mass} gets translated to \textit{density}.

\begin{definition}[Probability density function \textit{pdf}]
	For a continuous random variable x
	\[
	P(x:x\in A)= \int_{A} p_X(x)\dd x
	\]
	where $p_X(x)$ is the \textit{probability density function} of $x$ 
\end{definition}\todo{If we find a better definition, let's be sure to put it here. This one works, but it's not satisfying}

We use the term "function", but if we intend the integral in the Lebesgue sense we can extend the definition to \textit{pdf's} that are not functions but rather distributions. In general, this allows us a series of liberties that functions formally would not allow, such as mixing discrete and continuous variables. It's not guaranteed that a \textit{pdf} satisfies the Kolmogorov axioms, and the following example shows that while $p(x)\geq 0$ not necessarily $p(x) < 1$.

	\[
	p(x) = \begin{cases} 2 & \text{if } x \in [0,\frac{1}{2}] \\ 0 &  \text{otherwise} \end{cases}
	\]

However, it is crucial that $\int p(x)\dd x = P(X\in O_X) = 1$ (we say that the \textit{pdf} is \textit{normalized}), and one can verify that the example above satisfies this.\sidenote{Interesting to note as well that while probabilities in the Kolmogorov sense are pure dimensionless numbers, \textit{pdf's} have the dimensions of $[x]^{-N}$ where $N$ is the dimension of the space. This has natural parallels to the concept of mass and density in physics.} This normalization also brings us to define the \textbf{cumulative density function} of a distribution, as assures us that it is a monotonically increasing function of $x$ from $0$ to $1$.

\begin{definition}[Cumulative density function \textit{cdf}]
	For a continuous random variable x
	\[
	F_X(x)\equiv P\left(X\in \left(-\infty, x\right]\right) = \int_{-\infty}^{x}p_X(x^\prime)\dd x^\prime
	\]
\end{definition}

The fundamental theorem of calculus also tells us that $\dv{F}{x}=p(x)$. We will find the cumulative easier to calculate than the density function itself in many cases. 


Moving on to probability distributions for more than one random variable, things generalize quite easily. We can generally call $P(x_1, \allowbreak x_2, \allowbreak \dots, \allowbreak x_N)$ the probability distribution of a countable number of random variables, each defined on its own outcome space. We can define independence as we have in section \ref{subsec:kolmogorov}, and say that probability distributions (defined on the same $S$) are independent if 


	\[
	\forall A,B \quad P\left( \left( x\in A\right) \cap \left( y\in B\right) \right)  = P\left( x\in A\right) \cdot P\left( y\in B\right) 
	\]
	
\quad

This can also be written as $P_{XY}(x,y)=P_X(x)\cdot P_Y(y)$.\todo{This condition is necessary, but not sufficient, so when you have an example, put it here}



Much like the concept of independence, we generalize conditional probability and Bayes' theorem to distributions, obtaining
	\[
	P_X(x|y)\cdot P_Y(y) = P_{XY}(x,y) = P_Y(y|x)\cdot P_x(x)
	\]
	
Finally, to end our parallels with Kolmogorov probability we consider a partition $\{Y_i\}$ of $O_Y$, particularly one where $Y$ is discrete-valued such that $Y_i = \{y_i\}$. In this case we have
	\[
	P_X(x) = \sum_i P_{XY}(x,y_i)=\sum_i P_X(x|y_i)\cdot P_Y(y_i)
	\]
	
where astute readers will have identified the total probability law. This process is called \textit{marginalizing}, and $P_X(x)$ is the \textit{marginalized} distribution of $P_{XY}(x,y)$ with respect to $y$. This concept, although easy, has been known to be a "surprise question" at the exams that students tend to overlook.\sidenote{One of the authors speaks from experience.}

\subsection{The law of large numbers}
The law of large numbers (LLN) is one of the most fundamental and well-known theorems of this field. Let's start with a few definitions:
\begin{definition}[Arithmetic mean]
	Given a random variable $X$, we define \textbf{arithmetic mean} (or \textit{sample mean on N}) the statistic:
	\[
	\bar{x}_N = \frac{1}{N}S_N = \frac{1}{N}\sum_i^N x_i
	\]
\end{definition}

\begin{definition}[Mean]
	If $p_X(x)$ is the \textit{pdf} of $X$, we call the \textit{mean} of the distribution (or the expected value of $X$) the integral (Lebesgue):
	\[
	\langle x\rangle\equiv \E[x] \equiv \int_{O_X} x\cdot p_X(x)\dd x
	\]
\end{definition}

\begin{theorem}[Law of large numbers (LLN)]
	Let $X$ be a real-valued observable, and $p(x)$ its \textit{pdf}. Let's suppose that the mean of $X$ exists and is finite:
	\begin{equation}
	\mu = \E[x]=\int_{O_X} x\cdot p_X(x)\dd x
	\label{eq:LLN_mean}
	\end{equation}
	We then have that 
	\begin{equation}
	\forall \varepsilon > 0: \lim_{N\to\infty} P\left( \left| \bar{x}_N - \mu\right| > \varepsilon\right) =0
	\label{eq:LLN}
	\end{equation}
\end{theorem}

\quad

Within this theorem we have defined an interesting property "under the rug", which is convergence in probability, which is nothing more than the limit in (\ref{eq:LLN}).\todo{Maybe, instead of defining it this half-assed way we could add a definition explicitly?}

If we had to write the law of large numbers in words, it would read: "the sample mean of an observable converges in probability to its expected value whenever the latter exists and is finite". 
